### 2023-03-26 discussion.qmd
finalize typo check

In multiclass classification task, the goal is to predict the label (i.e. class) of a given input among a defined number of labels. In @sec-uvp6, I explained that XGBoost minimizes a regularized objective function $L$ (see @eq-objective-function-xgboost) during model training. In practice, the XGBoost library proposes two choices of objective functions for multiclass classification problems: `multi:softmax` and `multi:softprob` (https://xgboost.readthedocs.io/en/latest/parameter.html, last accessed: 03/03/2023). Both functions are based on the softmax normalization which normalizes the classification score of each class using an exponential function (@eq-softmax). `multi:softmax` returns the class of highest probability whereas `multi:softprob` outputs a class probability array, which allows the computation of custom objective functions.

@@

```{r, trust index, dpi = 300, fig.height = 12, fig.width= 10}
#| label: fig-trust-index
#| fig-cap: Percentage of error captured as a function of rank for four metrics (max, softmax, diff$_{max}$ and diff$_{softmax}$) on six UVP5 classes (4 biological, 2 non biological). The number of objects predicted in each class is provided in the title of each subplot.

@@

The second critical aspect of the model was its limitation at 40 classes. In the UVP5 SD data set, the most detailed (human) classification at a taxonomic level was roughly 150 classes (referred to as $level_{0}$). It was therefore necessary to merge classes based on morphological and ecological attributes to decrease that number in order to get to $level_{1}$ that had 38 classes (reminder, it is currently limited at 40). From $level_{1}$, we thus experimented a lot of different prediction models with a hyperparameters gridsearch in order to find both the right set of hyperparameters and the right number of classes. Still, some $level_{1}$ classes had very low precision and recall scores because they were confused with some specific taxonomic groups hence we merged them accordingly with those groups. This trial-error approach was necessary to create the final aggregation level ($level_{2}$) used for the final UVP5 model. The same work had to be done later with the UVP6 data because the taxonomic groups were not identical.

@@

@tbl-stats-uvp6 shows that the median number of objects detected in 3.5 m$^{3}$ is roughly 300. In addition, it is usual to have less than 20% of living objects hence the vast majority of images are labelled as detritus. The latter are ubiquitous in the ocean and display a wide variety of shapes, sizes and textures (@fig-detritus). In @sec-improve-classif, I showed that using several classes of detritus did not improve the classification model and that it actually made it worse because the N-detritus classes showed a high level of confusion between each other.

@@

![Examples of UVP objects labelled as detritus. The scale bar is the same for all (2 mm).](image/detritus/moz_detritus.jpg){#fig-detritus}

![Examples of UVP objects labelled as living material. The scale bar is the same for all (2 mm). Note that I have hidden three detritus from @fig-detritus to highlight the difficulty of classifying plankton already at the human level.](image/detritus/uvp_user_meeting/living_except3.jpg){#fig-living}

@@

[UVPec](https://github.com/ecotaxa/uvpec/) - short name for Underwater Vision Profiler Embedded Classifier - is a toolbox to train classification models for UVP6 images and evaluate their performance. This python package has been designed to be as user-friendly as possible in order to be accessible to the largest number of users. Basically, the user only needs to create folders (40 maximum because the maximum number of classes is fixed at 40) of labelled UVP6 images and specify some hyperparameters (e.g. learning rate) to train a classification model. The evaluation of the created model provides a confusion matrix and a classification report to help the user evaluate the quality of the model as well as the cross entropy curve to have information on the optimal boosting round number (before the model overfits).

@@

```{=latex}
\newpage
```
The observed concentrations of the 4 biological classes in @fig-kerguelen needs to be interpreted with the UVP6 classification report (@tbl-classification-report-uvp6-xgboost-cnn). *Calanoida* have a precision and recall of, respectively, 66% and 69% thus we potentially miss 31% of them (i.e. 31% of *Calanoida* are non detected because they are classified in another class) but 34% of images classified as *Calanoida* are false positives and should be placed in another class. In that case, the observed concentrations can therefore be resumed as the real concentration. *Chaetognatha* have a precision and recall of, respectively, 80% and 62% therefore when a *Chaetognaha* is detected, the model is confident (20% of false positives) but it misses 38% of them so we might have more *Chaetognatha* than measured by the UVP6. However, One must also look at the concentration values because it is likely that the 2 points measured on the eastern part of the Keguerlen Plateau are misclassified objects. We could therefore conclude to the absence *Chaetognatha* on the eastern part while their presence is likely on the western part because, on the one hand, this class has a high precision and, on the other hand, they are observed on successive short-lapsed vertical profiles and between 0-500 m where they likely feed on *Calanoida*.

@@

Unless we find a cheap and efficient (low power) way to send UVP6 images via Iridium communication, we are stuck with the issue of embedded classification. In @sec-uvp6, I showed that CNNs usually outperform features-based classifiers. However, the field of computer vision is constantly evolving with the arrival of new technologies that could help improve the automatic classification of zooplankton images. Transformers [@Vaswani2017-il], also referred to as Foundation Models in @Bommasani2021-kb are the latest cutting-edge tool in the field of computer vision and natural language processing [e.g. @Liu2021-jq; @Li2022-qt]. The structure of transformers relies on attention mechanisms able to draw global dependencies between input and output [@Vaswani2017-il]. In other words, it is able to learn dependencies between distant positions in a network architecture in constrast to CNNs. As an example, @Li2022-qt used a vision transformer model combining a CNN and a vision transformer to identify apple diseases. In their model, the CNN is used to extract the global features of the input image while the transformer extracts local features of the disease region to help the CNN see better. How does it help the CNN? In the case of apple images with different backgrounds (different weather conditions, brightness changes, other objects in the field of view), the CNN could be "distracted" hence the transformer is able to focus the attention of the CNN on the region of interest (the disease region) and not on irrelevant features such as the weather. The use of transformers for automatic zooplankton classification has just begun [@Kyathanahally2022-ma] but it should certainly be explored further in the near future.

@@

There is a class called "other\<living" in the UVP6 classification model. I did not mention earlier than we were forced to place copepods (except *Calanoida*) in that class. Indeed, *Calanoida* are a "large" species of copepods compared to other copepods found in the Mediterranean Sea that are smaller. The classification score of the latter was low (low precision and low recall) hence we put them in the "other\<living" class, which is a shame.

The good news is that the next generation of UVP is already in the making with the UVP6m (m for micro) which will be able to look at smaller objects (particle counter between 10 and 200 $\mu$m and plankton/particle imaging \> 100 $\mu$m). The UVP6m will be more compact than the UVP6 but the principles will be the same and both instruments will share 5 size classes for intercalibration.

@@

Before working on the CONVERSE framework, I was trying to find a global relationship (I also tried to find a regional one, without more success) to derive carbon fluxes from UVP5 PSDs. The idea was to reproduce the method of @Guidi2008-vp on the newly quality controlled UVP5 PSD data set [@Kiko2022-sk]. For this, I needed to build a global data set of POC measurements derived from sediment traps and $^{234}Th$ (see @tbl-poc-from-sediment-traps, @tbl-poc-from-thorium and [here](https://fricour.shinyapps.io/carbon_fluxes_app/) for a more interactive view). Here below, I briefly review three methods we explored to find a POC-PSD relationship and what could be further investigated.

@@

Instead of directly predicting carbon fluxes using environmental variable, we could focus on predicting PSD slope or PSD. In a recent paper, @Clements2022-oq used physical and biological variables (see their Table 1) to predict global maps of PSD slope and biovolume at the base of the EZ using UVP5 data and a random forest algorithm. @Clements2022-oq highlighted the importance of Chla (i.e. proxy of biomass), oxygen (i.e. proxy of respiration) and MLD as predictors in their random forest model. They also emphasized the importance of including phytoplankton functional groups and abundance as key controlling factors on both the PSD slope and the biovolume in future models due to their impact on the downward POC flux (i.e. aggregation/disaggregation processes). The study of @Denvil-Sommer2022-oi used both random forest and XGBoost models to predict small and large POC concentrations using geographical variables (depth, latitude and longitude), physical variables (incident light, MLD and temperature), chemical (phosphate and nitrate) and biological variables (Chla and 12 Plankton Functional Types (PFTs) such as diatoms or microzooplankton). @Denvil-Sommer2022-oi showed that for small POC, the most important predictors are the depth level, temperature, microzooplankton and phosphate while the large POC is mostly driven by depth, temperature, MLD, microzooplankton, phaeocystis (another PFT), phosphate and Chla averaged over the MLD. Overall, the results of these studies suggest that it is possible to link the surface environmental and ecosystem structure to the POC distribution in the ocean interior.

@@

The embedded classification for AUVs contributes to a better comprehension and quantification of the vertical migration pump. So far, BGC-Argo floats equipped with the UVP6 and the classification model have been deployed in some oceanic regions (e.g. Labrador Sea, Kerguelen Plateau) and have just started to acquire data. In addition to the particle (\> 0.6 mm) identification, the UVP6 also sends their mean area and gray levels which will hopefully contribute to better characterize their morphological traits and therefore better predict their sinking rates (cfr. size versus excess density). The latter are important to estimate the downward carbon flux and the sequestration time.

>>>>

In multiclass classification task, the goal is to predict the label (i.e. class) of a given input among a defined number of labels. In @sec-uvp6, I explained that XGBoost minimizes a regularized objective function $L$ (see @eq-objective-function-xgboost) during model training. In practice, the XGBoost library proposes two choices of objective functions for multiclass classification problems: `multi:softmax` and `multi:softprob` (https://xgboost.readthedocs.io/en/latest/parameter.html, last accessed: 03/03/2023). Both functions are based on the softmax normalization which normalizes the classification score of each class using an exponential function (@eq-softmax). `multi:softmax` returns the class of highest probability whereas `multi:softprob` returns a class probability array, which allows the computation of custom objective functions.

@@

```{r, trust index, dpi = 300, fig.height = 12, fig.width= 10}
#| label: fig-trust-index
#| fig-cap: Percentage of error captured as a function of rank for four metrics (max, softmax, diff$_{max}$ and diff$_{softmax}$) on six UVP5 classes (4 biological, 2 non-biological). The number of objects predicted in each class is provided in the title of each subplot.

@@

The second critical aspect of the model was its limitation at 40 classes. In the UVP5 SD data set, the most detailed (human) classification at a taxonomic level was roughly 150 classes (referred to as $level_{0}$). It was therefore necessary to merge classes based on morphological and ecological attributes to decrease that number in order to get to $level_{1}$ that had 38 classes (reminder, it is currently limited at 40). From $level_{1}$, we thus experimented a lot of different prediction models with a hyperparameters grid search in order to find both the right set of hyperparameters and the right number of classes. Still, some $level_{1}$ classes had very low precision and recall scores because they were confused with some specific taxonomic groups hence we merged them accordingly with those groups. This trial-error approach was necessary to create the final aggregation level ($level_{2}$) used for the final UVP5 model. The same work had to be done later with the UVP6 data because the taxonomic groups were not identical.

@@

@tbl-stats-uvp6 shows that the median number of objects detected in 3.5 m$^{3}$ is roughly 300. In addition, it is usual to have less than 20% of living objects hence the vast majority of images are labeled as detritus. The latter are ubiquitous in the ocean and display a wide variety of shapes, sizes and textures (@fig-detritus). In @sec-improve-classif, I showed that using several classes of detritus did not improve the classification model and that it actually made it worse because the different detritus classes showed a high level of confusion between each other.

@@

![Examples of UVP objects labeled as detritus. The scale bar is the same for all (2 mm).](image/detritus/moz_detritus.jpg){#fig-detritus}

![Examples of UVP objects labeled as living material. The scale bar is the same for all (2 mm). Note that I have hidden three detritus from @fig-detritus to highlight the difficulty of classifying plankton already at the human level.](image/detritus/uvp_user_meeting/living_except3.jpg){#fig-living}

@@

[UVPec](https://github.com/ecotaxa/uvpec/) - short name for Underwater Vision Profiler Embedded Classifier - is a toolbox to train classification models for UVP6 images and evaluate their performance. This python package has been designed to be as user-friendly as possible in order to be accessible to the largest number of users. Basically, the user only needs to create folders (40 maximum because the maximum number of classes is fixed at 40) of labeled UVP6 images and specify some hyperparameters (e.g. learning rate) to train a classification model. The evaluation of the created model provides a confusion matrix and a classification report to help the user evaluate the quality of the model as well as the cross entropy curve to have information on the optimal boosting round number (before the model overfits).

@@

```{=latex}
\newpage
```
The observed concentrations of the 4 biological classes in @fig-kerguelen needs to be interpreted with the UVP6 classification report (@tbl-classification-report-uvp6-xgboost-cnn). *Calanoida* have a precision and recall of, respectively, 66% and 69% thus we potentially miss 31% of them (i.e. 31% of *Calanoida* are non detected because they are classified in another class) but 34% of images classified as *Calanoida* are false positives and should be placed in another class. In that case, the observed concentrations can therefore be assumed as the real concentration. *Chaetognatha* have a precision and recall of, respectively, 80% and 62% therefore when a *Chaetognatha* is detected, the model is confident (20% of false positives) but it misses 38% of them so we might have more *Chaetognatha* than measured by the UVP6. However, One must also look at the concentration values because it is likely that the 2 points measured on the eastern part of the Kerguelen Plateau are misclassified objects. We could therefore conclude to the absence of *Chaetognatha* on the eastern part while their presence is likely on the western part because, on the one hand, this class has a high precision and, on the other hand, they are observed on successive short-lapsed vertical profiles and between 0-500 m where they likely feed on *Calanoida*.

@@

Unless we find a cheap and efficient (low power) way to send UVP6 images via Iridium communication, we are stuck with the issue of embedded classification. In @sec-uvp6, I showed that CNNs usually outperform features-based classifiers. However, the field of computer vision is constantly evolving with the arrival of new technologies that could help improve the automatic classification of zooplankton images. Transformers [@Vaswani2017-il], also referred to as Foundation Models in @Bommasani2021-kb are the latest cutting-edge tools in the field of computer vision and natural language processing [e.g. @Liu2021-jq; @Li2022-qt]. The structure of transformers relies on attention mechanisms able to draw global dependencies between input and output [@Vaswani2017-il]. In other words, it is able to learn dependencies between distant positions in a network architecture in constrast to CNNs. As an example, @Li2022-qt used a vision transformer model combining a CNN and a vision transformer to identify apple diseases. In their model, the CNN is used to extract the global features of the input image while the transformer extracts local features of the disease region to help the CNN see better. How does it help the CNN? In the case of apple images with different backgrounds (different weather conditions, brightness changes, other objects in the field of view), the CNN could be "distracted" hence the transformer is able to focus the attention of the CNN on the region of interest (the disease region) and not on irrelevant features such as the weather. The use of transformers for automatic zooplankton classification has just begun [@Kyathanahally2022-ma] but it should certainly be explored further in the near future.

@@

There is a class called "other\<living" in the UVP6 classification model. I did not mention earlier that we were forced to place copepods (except *Calanoida*) in that class. Indeed, *Calanoida* are a "large" species of copepods compared to other copepods found in the Mediterranean Sea that are smaller. The classification score of the latter was low (low precision and low recall) hence we put them in the "other\<living" class, which is a shame.

The good news is that the next generation of UVPs is already in the making with the UVP6m (m for micro) which will be able to look at smaller objects (particle counter between 10 and 200 $\mu$m and plankton/particle imaging \> 100 $\mu$m). The UVP6m will be more compact than the UVP6 but the principles will be the same and both instruments will share 5 size classes for intercalibration.

@@

Before working on the CONVERSE framework, I was trying to find a global relationship (I also tried to find a regional one, without more success) to derive carbon fluxes from UVP5 PSDs. The idea was to reproduce the method of @Guidi2008-vp on the newly quality controlled UVP5 PSD data set of @Kiko2022-sk. For this, I needed to build a global data set of POC measurements derived from sediment traps and $^{234}Th$ (see @tbl-poc-from-sediment-traps, @tbl-poc-from-thorium and [here](https://fricour.shinyapps.io/carbon_fluxes_app/) for a more interactive view). Here below, I briefly review three methods we explored to find a POC-PSD relationship and what could be further investigated.

@@

Instead of directly predicting carbon fluxes using environmental variables, we could focus on predicting PSD slopes or PSDs. In a recent paper, @Clements2022-oq used physical and biological variables (see their Table 1) to predict global maps of PSD slope and biovolume at the base of the EZ using UVP5 data and a random forest algorithm. @Clements2022-oq highlighted the importance of Chla (i.e. proxy of biomass), oxygen (i.e. proxy of respiration) and MLD as predictors in their random forest model. They also emphasized the importance of including phytoplankton functional groups and abundance as key controlling factors on both the PSD slope and the biovolume in future models due to their impact on the downward POC flux (i.e. aggregation/disaggregation processes). The study of @Denvil-Sommer2022-oi used both random forest and XGBoost models to predict small and large POC concentrations using geographical (depth, latitude and longitude), physical (incident light, MLD and temperature), chemical (phosphate and nitrate) and biological variables (Chla and 12 Plankton Functional Types (PFTs) such as diatoms or microzooplankton). @Denvil-Sommer2022-oi showed that for small POC, the most important predictors are the depth level, temperature, microzooplankton and phosphate while the large POC is mostly driven by depth, temperature, MLD, microzooplankton, phaeocystis (another PFT), phosphate and Chla averaged over the MLD. Overall, the results of these studies suggest that it is possible to link the surface environmental and ecosystem structure to the POC distribution in the ocean interior.

@@

The embedded classification for AUVs will contribute to a better comprehension and quantification of the vertical migration pump. So far, BGC-Argo floats equipped with the UVP6 and the classification model have been deployed in some oceanic regions (e.g. Labrador Sea, Kerguelen Plateau) and have just started to acquire data. In addition to the particle (\> 0.6 mm) identification, the UVP6 also sends their mean area and gray levels which will hopefully contribute to better characterize their morphological traits and therefore better predict their sinking rates (cfr. size versus excess density debate). The latter are important to estimate the downward carbon flux and the sequestration time.

### 2023-03-26 index.qmd
finalize typo check

# Foreword {.unnumbered}

As of 2022, it has now become obvious to the general public that the Earth's climate is changing. Massive flood events, repeated heatwaves, never-ending droughts, disappearing glaciers, biodiversity collapse, … Reports on weather anomalies were already published in the 1960's, the Intergovernmental Panel on Climate Change (IPCC) was created in 1988, the Kyoto Protocol was adopted in 1997, An Inconvenient Truth was released in 2006, the Paris Agreement was signed in 2015 and still, what has really changed?

@@

Although I am not an activist (yet?) or a politician (I am too honest to be elected), rest assured that there is no fake news in this work. I hope you will enjoy it.

>>>>

# Summary {.unnumbered}

**Towards a new insight of the carbon transport in the global ocean**

The ocean is known to play a key role in the carbon cycle. Without it, atmospheric CO$_{2}$ levels would be much higher than what they are today thanks to the presence of carbon pumps that maintain a gradient of dissolved inorganic carbon (DIC) between the surface and the deep ocean. The biological carbon pump (BCP) is primarily responsible for this gradient. It consists in a series of ocean processes through which inorganic carbon is fixed as organic matter by photosynthesis in sunlit surface waters and then transported to the ocean interior and possibly the sediment where it will be sequestered from the atmosphere for millions of years. The BCP was long thought as solely the gravitational settling of particulate organic carbon (POC). However, a new paradigm for the BCP has recently been defined in which physically and biologically mediated particle injection pumps have been added to the original definition. Physically mediated particle injection pumps provide a pathway to better understand the transport of dissolved organic carbon (DOC) whereas biologically mediated particle injection pumps focus on the transport of POC by vertically migrating animals, either daily or seasonally. Therefore, a better understanding of these processes could help bridge the gap between carbon leaving the surface and carbon demand in the ocean interior. To address this new paradigm, this work will benefit from the advent of recent sensors that equip a new generation of Biogeochemical-Argo floats (BGC-Argo). The first part focuses on the development of an embedded zooplankton classification model for the Underwater Vision Profiler 6 (UVP6) under strict technical and energy constraints. The second part studies particle and carbon fluxes in the Labrador Sea using BGC-Argo floats equipped for the first time with the UVP6 and an optical sediment trap (OST), providing two independent measurements of sinking particles. The last part consists in revisiting the BCP using a new framework called CONVERSE for Continuous Vertical Sequestration. With this new approach, we re-evaluate the total carbon sequestered from the atmosphere (\> 100 years) by the BCP and its transport pathways on the entire water column, in contrast to the carbon sequestration typically assumed below a fixed reference depth.

**Keywords** Carbon pumps -- Underwater Vision Profiler 6 -- BGC-Argo floats -- Machine learning

```{=latex}
\newpage
```
**Vers une meilleur compréhension du transport du carbone dans l'océan global**

L'océan est connu pour jouer un rôle clé dans le cycle du carbone. Sans lui, les niveaux de CO$_{2}$ atmosphérique seraient bien plus élevés qu'aujourd'hui grâce à la présence de pompes à carbone qui maintiennent un gradient de carbone inorganique dissous (DIC) entre la surface et le fond de l'océan. La pompe à carbone biologique (BCP) est principalement responsable de ce gradient. Elle consiste en une série de processus océaniques au cours desquels le carbone inorganique est converti en matière organique via la photosynthèse dans les eaux de surface, puis transporté vers l'intérieur de l'océan et éventuellement les sédiments où il sera séquestré par rapport à l'atmosphère pour des millions d'années. La BCP était longtemps considérée comme étant uniquement la déposition gravitationnelle de carbone organique particulaire (POC). Cependant, un nouveau paradigme pour la BCP a récemment été défini dans lequel des pompes d'origine physique et biologique d'injection de particules ont été ajoutées à la définition originale. Les pompes physiques d'injection de particules fournissent un moyen de mieux comprendre le transport de carbone organique dissous (DOC), tandis que les pompes biologique d'injection de particules se concentrent sur le transport de POC par des animaux migrant verticalement, quotidiennement ou saisonnièrement. Par conséquent, une meilleure compréhension de ces processus pourrait aider à combler l'écart entre le carbone quittant la surface et la demande de carbone dans l'océan profond. Pour aborder ce nouveau paradigme, ce travail bénéficiera de l'arrivée de capteurs récents équipant une nouvelle génération de flotteurs Biogéochimiques-Argo (BGC-Argo). La première partie se concentre sur le développement d'un modèle embarqué de classification de zooplancton pour l'Underwater Vision Profiler 6 (UVP6) avec des contraintes techniques et énergétiques strictes. La deuxième partie étudie les flux de particules et de carbone dans la mer du Labrador en utilisant des flotteurs BGC-Argo équipés pour la première fois de l'UVP6 et d'un piège à sédiments optique (OST), fournissant deux mesures indépendantes des particules. La dernière partie consiste à revisiter la BCP en utilisant un nouveau cadre appelé CONVERSE qui fait référence à la séquestration verticale continue du carbone dans la colonne d'eau. Avec cette nouvelle approche, nous réévaluons le carbone total séquestré par rapport à l'atmosphère (\> 100 ans) par la BCP et ses voies de transport sur toute la colonne d'eau, par opposition à la séquestration de carbone généralement supposée en-dessous d'une profondeur de référence fixe.

**Mots clés** Pompes à carbone -- Underwater Vision Profiler 6 -- Flotteurs BGC-Argo -- Machine learning

```{=latex}
\newpage
```

<!-- # Foreword {.unnumbered}

As of 2022, it has now become obvious to the general public that the Earth's climate is changing. Massive flood events, repeated heatwaves, never-ending droughts, disappearing glaciers, biodiversity collapse, ... Reports on weather anomalies were already published in the 1960's, the Intergovernmental Panel on Climate Change (IPCC) was created in 1988, the Kyoto Protocol was adopted in 1997, An Inconvenient Truth was released in 2006, the Paris Agreement was signed in 2015 and still, what has really changed?

@@

Although I am not an activist (yet?) or a politician (I am too honest to be elected), rest assured that there is no fake news in this work. I hope you will enjoy it. -->

### 2023-03-26 merci.qmd
typo check
finalize typo check

# Acknowledgments

Thanks to the members of the jury for taking the time to read and evaluate this three and a half year work. I am looking forward to debate it during the defense.

Cette thèse a été particulière à bien des égards, avec son lot de positifs et de négatifs mais surtout avec son lot de rebondissements qui m'ont permis d'affûter mes connaissances, ma résilience et je l'espère ma tolérance. Pour cela, je tiens à remercier chaleureusement toutes les personnes y ayant contribué de près ou de loin.

@@

Merci également aux personnes que j'ai pu cotoyer à COMPLEX comme chez OMTAB depuis mon passage en stage en 2018 et les multiples allers-retours Belgique-France.

@@

Merci aux Montois (Mathieu, Chris, Flo, Ju & Ju) pour votre soutien à distance, un mariage en 3 actes et pour vos visites sur Nice. Petite dédicace à Ju K qui, à force d'avoir joué avec le diable en subissant l'enfer des pourcentages, va littéralement en devenir un. Vivement 2024 pour casser la baraque.

>>>>

# Acknowledgments {.unnumbered}

I would like to thank the members of the jury for taking the time to read and evaluate this three and a half year work. I look forward to discussing it during the defense.

Merci aux membres du comité de thèse pour leur temps et leurs retours sur l'évolution du projet.

Cette thèse a été particulière à bien des égards, avec son lot de points positifs et négatifs mais surtout avec son lot de rebondissements qui m'ont permis d'affûter mes connaissances, ma résilience et je l'espère ma tolérance. Pour cela, je tiens à remercier chaleureusement toutes les personnes y ayant contribué de près ou de loin.

@@

Merci également aux personnes que j'ai pu côtoyer à COMPLEX comme chez OMTAB depuis mon passage en stage en 2018 et les multiples allers-retours Belgique-France.

@@

Merci aux Montois (Mathieu, Chris, Flo, Ju & Ju) pour votre soutien à distance, un mariage en 3 actes et pour vos visites sur Nice. Petite dédicace à Ju K qui, à force d'avoir joué avec le diable en subissant l'enfer des pourcentages, va littéralement en devenir un. Vivement 2024 pour casser la baraque!
